---
title: "Data Profile - World Health Organization API"
author: "Kurt Wydrinski"
date: "11/22/2019"
output: pdf_document
---

```{r setup, include=FALSE}
# include source chunks in knitted output by default
knitr::opts_chunk$set(echo = TRUE)
# include the necessary packages
library(tidyverse)
```

# Source

During the formation of the United Nations in 1945, one of the key objectives
for its ongoing mission was to foster cooperation between countries to address
many social well-being issues, including health.[^1] This objective was taken on
as part of the World Health Organization (WHO) which was established on April 7,
1948.[^2] The WHO continues this mission today with 194 countries cooperating at
various levels to promote global health concerns. In doing so, the WHO regularly
collects data from its members that are related to health issues. The data
contains various time series regarding diseases, illnesses, economics, social
demographics, etc. This data is collectively captured in the WHO's online
database *Global Health Observatory (GHO)*.[^3]

[^1]: See <https://www.history.com/this-day-in-history/the-united-nations-is-born>.
[^2]: See <https://www.who.int/about/who-we-are/history>.
[^3]: See <https://www.who.int/gho/about/en/>.

# API Overview

The R package `WHO`, provides a simple API to access the GHO. It only provides
two functions: `get_codes()` and `get_data()`. Inside the GHO, each time series
that exists is identifed by a label. Each label is a code that uniquely
identifies the series. These labels are then used a a parameter to `get_data()`
to retrieve the time series observations.

```{r}
# install.packages("WHO")
library(WHO)
```

The code below uses the `extra` parameter to download all metadata available for
the GHO codes.

```{r}
who_codes <- get_codes(extra = TRUE)
glimpse(who_codes)
```

There are `r formatC(length(who_codes$label), format = "d", big.mark = ",")`
indicators or time series in this dataset. The `category` variable is a WHO
grouping of the indicators into `r n_distinct(who_codes$category)` sets of
indicators as follows:

\tiny

```{r}
who_codes %>%
  group_by(category) %>%
  summarise(indicator_count = n()) %>%
  knitr::kable()
```

\normalsize

Note the following about the above categories.

1. There are 447 indicators that do not have a category assigned by the WHO.
   These should be considered as a category called `Uncategorized` or
   `No category`.
2. The naming convention is not consistent across categories. For example, some
   use single words, phrases, all capital letters, irregular case, etc. Category
   names should be somewhat standardized while retaining the original meaning in
   the dataset.
3. A number of categories appear to be fragmented. For example, there are
   three premutations of *Neglected Tropical Diseases* that need to be combined
   into a single category.

The table below provides a list of the indicators in the
*Demographic and socioeconomic statistics* category. However, inspection of the
table should make it apparent that a number of these indicators are based on
population.

```{r}
who_codes %>%
  filter(category == "Demographic and socioeconomic statistics") %>%
  select(label, display) %>%
  knitr::kable()
```

A closer look may make it apparent that the data may not have a total population
number. It is possible that `WHS10_1` could have it by the `display` string.

```{r}
who_codes %>%
  filter(label == "WHS10_1")
```

Inspection of the indicator observation in the metadata shows that the `url` and
`definition_xml` variables are URLs to further information about the indicator.
`url` is an address for a web page on the GHO registry that explains what the
indicator represents. The `defintion_xml` provides the same information
presented in the explanation page in XML format.

As the [explanation page](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/3056)
suggests, the `WHS10_1` indicator does reflect the population count/census
statistic. This may or may not be useful when used with other indicators. In
looking at all indicators in the metadata, there are at least three others that
are called `Population`.

```{r}
who_codes %>%
  filter(str_detect(display, "^Population$")) %>%
  select(label, display, category)
```

## API Risks

Based on the analysis above, the following risks appear to exist when using this
API dataset.

1. The data is not tidy and needs to be made such before detailed analysis can
   be completed.
2. The structure of the data is inconsistent and needs detailed exploration when
   deciding on which indicators to use.
3. The `category` variable in the metadata (i.e. codes) is not very reliable.
   Careful considerations need to be made when leveraging this variable. The
   category should be used as the basis for a new variable which is a reliable
   identifier for logically-related indicators.

## API Rewards

Based on the analysis above the following benefits seem to be gained by using
this API and dataset.

1. The data can be accessed using the simple `WHO` package.
2. The data is acessible via the Internet via on-demand API calls.
3. The data contains international health and disease data along with related
   indicators.

# Usage Prototype

The sections below provide a prototypical usage for this dataset and API. The
actual use and visualization of the data will vary between projects. However,
the sections below demonstrate common usage patterns for this data and API.

## Prototype Exploration

### *per 100,000* Indicators

A number of the indicators  in the database are measured and scaled to units of
100,000. Most these indicators are scaling to report over country population
units of 100,000 people. The table below presents a sampling of these
indicators.

```{r}
who_codes %>%
  filter(str_detect(display, "per 100 000 ")) %>%
  select(label, display, category)
```

#### Usefulness Assessment

There are about ninety indicators that report statistics on groups of 100,000.
Most of the grouping are by population (i.e. *per 100,000 people*) but there are
also a few by *adult population* and *live births*. As can be seen in the table
below, these indicators cross a number of topics. It could be possible to tell
a story about *The X per 100,000 people...*.

```{r echo=FALSE}
who_codes %>%
  filter(str_detect(display, "per 100 000 ")) %>%
  mutate(Topic = category) %>%
  group_by(Topic) %>%
  summarise(Indicators = n()) %>%
  arrange(Topic) %>%
  knitr::kable()
```

### *Infectious diseases* Indicators

A number of the indicators in the database measure various statistics about
different infectious diseases. There are statistics for number of cases, number
of deaths from various infectious diseases, and a few other more detailed
statsitics about diseases such as leptospirosis (not to be confused with
leprosy.) The table below presents a sampling of these indicators.

```{r}
who_codes %>%
  filter(category == "Infectious diseases")
```

Each of these indicators has a number of time series with annual values for
different countries. However, not all of these indicators have a lot of data and
may not be very usable. The sections below describe a few indicators that are of
interest and could be of some use.

#### "The Bird Flu" - Number of Cases

The `WHS3_51` indicator shows the *H5N1 influenza - number of reported cases*.
The H5N1 strain of influenza is commonly know as *avian flu* or *the bird flu*.

```{r}
# This function is used to create clean factors for analysis from the GHO
# dataset.
gho_vectorize <- function(v) {
  switch(class(v),
         "character" = {
           f <- as_factor(v)
           fct_explicit_na(fct_relevel(f,
                                       sort(levels(f))),
                           na_level = "NA")
         },
         "factor" = {
           fct_explicit_na(fct_relevel(v,
                                       sort(levels(v))),
                           na_level = "NA")
         },
         "numeric" = as_factor(v))
}
```

```{r}
tb_bird_flu <- get_data("WHS3_51")
tb_bird_flu$country <- gho_vectorize(tb_bird_flu$country)
tb_bird_flu$region <- gho_vectorize(tb_bird_flu$region)
tb_bird_flu$publishstate <- gho_vectorize(tb_bird_flu$publishstate)
```

```{r}
tb_bird_flu %>%
  group_by(country) %>%
  select(country, year, value, worldbankincomegroup, region) %>%
  arrange(country, year)
```

```{r}
summary(tb_bird_flu)
```

##### Usefulness Assessment

The following items may cause issues when using this indicator.

1. There is are only 28 observations for Bird Flu in the dataset.
2. There are 3 observations with `NA` for the country.
   2.1. Two of these are for the *Western Pacific* region which also includes
        Viet Nam, China, etc. These oberservations could be reports of incidents
        that could not be attributed to a given country in the region but may
        have been reported by one. For example, a foreign traveller could have
        been treated for the incident in Viet Nam. It is not fair to assume the
        person contracted the disease in Viet Nam so the `NA` country should be
        considered a valid observation.
   2.2. The third of these is attributed to the *Upper-middle-income* WHO income
        group category. No other observations have this attribution so this one
        appears to be an observation caputured by the WHO to highlight the group.

Based on the issues above, there is some elevated risk in using this dataset. It
is very small so statistics will be hard to accurately derive. The `NA` country
observations suggest a mixture of hetergeneous observations; most are
observations of the number of cases in a given country for a given year while
others summarize case reports for various groups that span countries or outside
of countries in some way.

#### Tuberculosis - Number of Cases

The `WHS3_54` indicator shows the *Number of reported cases of tuberculosis (DOTS)*.

```{r}
tb_tuber <- get_data("WHS3_54")
tb_tuber$country <- gho_vectorize(tb_tuber$country)
tb_tuber$region <- gho_vectorize(tb_tuber$region)
tb_tuber$publishstate <- gho_vectorize(tb_tuber$publishstate)
```

```{r}
tb_tuber %>%
  group_by(country) %>%
  arrange(country, year) %>%
  select(country, year, value, worldbankincomegroup, region)
```

```{r}
summary(tb_tuber)
```

```{r}
who_codes %>%
  filter(label == "WHS3_54")
```

##### Usefulness Assessment

The following item may cause issues when using this indicator.

1. There are only observations for the year 2008.

Based on the issue above, this data will have limited use bacause it only covers
one year.

#### Cholera - Number of Cases

The `CHOLERA_0000000001` indicator shows the
*Number of reported cases of cholera*.

```{r}
tb_cholera <- get_data("CHOLERA_0000000001")
tb_cholera$country <- gho_vectorize(tb_cholera$country)
tb_cholera$region <- gho_vectorize(tb_cholera$region)
tb_cholera$publishstate <- gho_vectorize(tb_cholera$publishstate)
```

```{r}
tb_cholera %>%
  group_by(country) %>%
  arrange(country, year) %>%
  select(country, year, value, worldbankincomegroup, region)
```

```{r}
summary(tb_cholera)
```

```{r, fig.width=6.5, fig.height=10}
ggplot(data = tb_cholera %>%
                group_by(country) %>%
                summarise(count = n())) +
  geom_col(mapping = aes(x = country,
                         y = count)) +
  coord_flip() +
  theme(axis.text = element_text(size = 6)) +
  labs(title = "Per Country Cholera Observations",
       subtitle = "(the number of annual observations about cholera)",
       caption = "Data retrieved via WHO::get_data('CHOLERA_0000000001').",
       tag = "GHO",
       x = "",
       y = "Observations")  
```

```{r}
tb_cholera %>%
  filter(country == "NA")
```

##### Usefulness Assessment

The following items may cause issues when using this indicator.

1. Not all countries have a cosnsistent number of annual observations.
2. There are 9 observations where `country` is `NA`.
   2.1. These were all from 2013.
   2.2. They appear to be summmary data produced in 2013 for different regions
        (including a "global" region).

Although the above issues pose some elevated risk in using this dataset, there
are enough observations across countries and years to analyze and report on this
data.

### *water* Indicators

Cholera is caused by the *Vibrio cholerae* bacterium which has infected either
food or water that has been ingested.[^4] The *GHO* has a number of indicators
in its datasets that could be useful to help tell a story about cholera. The
table below presents a sampling of these indicators.

[^4]: See <https://www.webmd.com/a-to-z-guides/cholera-faq#1>.

```{r}
who_codes %>%
  filter(str_detect(display, "water")) %>%
  select(category, display, label) %>%
  arrange(category, label)
```

#### Water - Improved Drinking Sources (`WHS5_122`)

The `WHS5_122` indicator shows the
*Population using improved drinking-water sources (%)*.

```{r}
tb_water_improved <- get_data("WHS5_122")
```

##### Usefulness Assessment

The following items will cause issues when using this indicator.

1. There are no observations available for this indicator.

Without any data, there is nothing to analyze or report. This indicator is no
useful.

#### Water - Improved Drinking Sources (`EQ_WATER`)

The `EQ_WATER` indicator shows the
*Population using improved drinking-water sources (%)*.

```{r}
tb_water_improved <- get_data("EQ_WATER")
tb_water_improved$country <- gho_vectorize(tb_water_improved$country)
tb_water_improved$region <- gho_vectorize(tb_water_improved$region)
tb_water_improved$publishstate <- gho_vectorize(tb_water_improved$publishstate)
```

```{r}
tb_water_improved %>%
  arrange(country, year) %>%
  select(country, year, value, wealthquintile, region)
```

```{r}
summary(tb_water_improved)
```

```{r, fig.width=6.5, fig.height=3}
ggplot(data = tb_water_improved %>%
                group_by(country) %>%
                summarise(count = n())) +
  geom_col(mapping = aes(x = country,
                         y = count)) +
  scale_y_continuous(breaks = seq(0, 15, 1)) +
  coord_flip() +
  theme(axis.text = element_text(size = 6)) +
  labs(title = "Per Country Improved Drinking Water Observations",
       subtitle = "(the number of annual observations about improved drinking water access)",
       caption = "Data retrieved via WHO::get_data('EQ_WATER').",
       tag = "GHO",
       x = "",
       y = "Observations")  
```

```{r}
ggplot(data = tb_water_improved %>%
                group_by(year) %>%
                summarise(count = n())) +
  geom_col(mapping = aes(x = year,
                         y = count)) +
  labs(title = "Annual Improved Drinking Water Observations",
       subtitle = "(the total annual observations about improved drinking water access)",
       caption = "Data retrieved via WHO::get_data('EQ_WATER').",
       tag = "GHO",
       x = "",
       y = "Observations")  

```

##### Usefulness Assessment

The following items may cause issues when using this indicator.

1. There are no observations for the years 1998-2002.
2. The dataset seems like it could be incomplete data. The WHO defintion of this
   indicator can be located at [the WHO web site](https://www.who.int/data/gho/data/indicators/indicator-details/GHO/population-using-improved-drinking-water-sources-(-)).
   At the bottom of the page, there is [a link](https://washdata.org/) to a
   joint project between WHO and UNICEF regarding safe drinking water. This web
   site contains data for more countries and for more years regarding safe
   drinking water. It appears the `WHO` package API may not be pulling all
   available data.
3. The data is not tidy because each year has 5 rows with each row represent the
   value for a certain quintile of the wealth of the population.

Based on the issue above, this data may have limited use bacause it only covers
a subset of the available data. The data at the joint project between the WHO
and UNICEF may be better to use than this dataset.

## Prototype Requirements

Although there are some limitations to this data, there should be enough usable
data to create some interested visualizations. If this data was used in
conjunction with the WHO/UNICEF data, we could tell a story with the following
visualizations.

1. Overview of different infectious diseases around the world.
2. Highlight of cholera incidents and deaths around the world.
3. Breakdown of water access by wealth quintiles in cholera-affected nations.
4. TBD more detail regarding wealth (or aid) with data TBD.

We'd need further analysis but should be able to easily find data to further
breakdown the wealth aspect. There many other indicator in the WHO data to
analyze and possibel use for this. If they aren't usuable, we can get economic
data regarding wealth elsewhere.

