---
title: "Data Profile - World Health Organization API"
author: "Kurt Wydrinski"
date: "11/22/2019"
output: pdf_document
---

```{r setup, include=FALSE}
# include source chunks in knitted output by default
knitr::opts_chunk$set(echo = TRUE)
# include the necessary packages
library(tidyverse)
```

# Source

During the formation of the United Nations in 1945, one of the key objectives
for its ongoing mission was to foster cooperation between countries to address
many social well-being issues, including health.[^1] This objective was taken on
as part of the World Health Organization (WHO) which was established on April 7,
1948.[^2] The WHO continues this mission today with 194 countries cooperating at
various levels to promote global health concerns. In doing so, the WHO regularly
collects data from its members that are related to health issues. The data
contains various time series regarding diseases, illnesses, economics, social
demographics, etc. This data is collectively captured in the WHO's online
database *Global Health Observatory (GHO)*.[^3]

[^1]: See <https://www.history.com/this-day-in-history/the-united-nations-is-born>.
[^2]: See <https://www.who.int/about/who-we-are/history>.
[^3]: See <https://www.who.int/gho/about/en/>.

# API Overview

The R package `WHO`, provides a simple API to access the GHO. It only provides
two functions: `get_codes()` and `get_data()`. Inside the GHO, each time series
that exists is identifed by a label. Each label is a code that uniquely
identifies the series. These labels are then used a a parameter to `get_data()`
to retrieve the time series observations.

```{r}
# install.packages("WHO")
library(WHO)
```

The code below uses the `extra` parameter to download all metadata available for
the GHO codes.

```{r}
who_codes <- get_codes(extra = TRUE)
glimpse(who_codes)
```

There are `r formatC(length(who_codes$label), format = "d", big.mark = ",")`
indicators or time series in this dataset. The `category` variable is a WHO
grouping of the indicators into `r n_distinct(who_codes$category)` sets of
indicators as follows:

\tiny

```{r}
who_codes %>%
  group_by(category) %>%
  summarise(indicator_count = n()) %>%
  knitr::kable()
```

\normalsize

Note the following about the above categories.

1. There are 447 indicators that do not have a category assigned by the WHO.
   These should be considered as a category called `Uncategorized` or
   `No category`.
2. The naming convention is not consistent across categories. For example, some
   use single words, phrases, all capital letters, irregular case, etc. Category
   names should be somewhat standardized while retaining the original meaning in
   the dataset.
3. A number of categories appear to be fragmented. For example, there are
   three premutations of *Neglected Tropical Diseases* that need to be combined
   into a single category.

The table below provides a list of the indicators in the
*Demographic and socioeconomic statistics* category. However, inspection of the
table should make it apparent that a number of these indicators are based on
population.

```{r}
who_codes %>%
  filter(category == "Demographic and socioeconomic statistics") %>%
  select(label, display) %>%
  knitr::kable()
```

A closer look may make it apparent that the data may not have a total population
number. It is possible that `WHS10_1` could have it by the `display` string.

```{r}
who_codes %>%
  filter(label == "WHS10_1")
```

Inspection of the indicator observation in the metadata shows that the `url` and
`definition_xml` variables are URLs to further information about the indicator.
`url` is an address for a web page on the GHO registry that explains what the
indicator represents. The `defintion_xml` provides the same information
presented in the explanation page in XML format.

As the [explanation page](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/3056)
suggests, the `WHS10_1` indicator does reflect the population count/census
statistic. This may or may not be useful when used with other indicators. In
looking at all indicators in the metadata, there are at least three others that
are called `Population`.

```{r}
who_codes %>%
  filter(str_detect(display, "^Population$")) %>%
  select(label, display, category)
```

## API Risks

Based on the analysis above, the following risks appear to exist when using this
API dataset.

1. The data is not tidy and needs to be made such before detailed analysis can
   be completed.
2. The structure of the data is inconsistent and needs detailed exploration when
   deciding on which indicators to use.
3. The `category` variable in the metadata (i.e. codes) is not very reliable.
   Careful considerations need to be made when leveraging this variable. The
   category should be used as the basis for a new variable which is a reliable
   identifier for logically-related indicators.

## API Rewards

Based on the analysis above the following benefits seem to be gained by using
this API and dataset.

1. The data can be accessed using the simple `WHO` package.
2. The data is acessible via the Internet via on-demand API calls.
3. The data contains international health and disease data along with related
   indicators.

# Usage Prototype

The sections below provide a prototypical usage for this dataset and API. The
actual use and visualization of the data will vary between projects. However,
the sections below demonstrate common usage patterns for this data and API.

## Prototype Assumptions / Requirements




