---
title: "Geographic Mapping Techniques"
author:
- Jedidiah Harwood^[California State University - East Bay]
- Kurt Wydrinski^[California State University - East Bay]
date: "11/22/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(maps)
```

# Introduction

There are a number of techniques available for creating geographic maps within
R. Techniques using both the `maps`, `ggplot` and other `tidyverse` packages are
demonstrated in the following sections.

# Basic Techniques

## World Maps

The code below demonstrates the simplest world map. By default, borders are
drawn 1 unit thick in a medium grey color. Also notice that the map is centered
on the Prime Meridian at $0^{\circ}$ longitude.

```{r}
ggplot() +
  borders()
```

The simple map is just that, simple but it's not very appealing. Adding the
`size` graphic aesthetic parameter to the borders call, thinner border line can
be used in which improves the appeal of the map.

```{r}
ggplot() +
  borders(size = 0.1)    # thin borders
```

While the thin borders are an improvement, they become harder to see due to the
default medium-grey color. Using the `colour` parameter of `borders()` allows
the borders to be rendered in different colors. The code below demonstrates
using a thin, black border. This fine-line border looks cleaner and more refined
than the default border rendering.

```{r}
ggplot() +
  borders(size = 0.1,
          colour = "black")    # thin, black borders
```

The axes, titles, captions, etc. are either missing or not labelled very clearly
in the above plot. The following actions will address this:

1. Re-scale axes and label with proper units-of-measure.
2. Clearly label variables represented on the axes.
3. Add placeholder titling elements for further clarity. However, these are
   placeholders because it is expected that each world map data visualization
   will be appropriately titled for the information it is conveying. These
   guidelines should be followed when choosing appropriate titling elements:
   
   1. `title` - Indicate the main subject/point of data visualization (e.g.
      *Country Population*)
   2. `subtitle` - Indicate constraints, filters or transformations applied to
      the underlying data (e.g. *(by most recent reported census)*, *(in thousands of people)*,
      etc.)
   3. `caption` - **TODO: define standard**
   4. `tag` - **TODO: define standard**

The code below applies to actions to improve the ease in understanding the map.

```{r}
ggplot() +
  borders(size = 0.1,
          colour = "black") +    # thin, black borders
  scale_x_continuous(breaks = seq(-180, 180, 90),
                     minor_breaks = seq(-180, 180, 30),
                     labels = unlist(
                       lapply(seq(-180, 180, 90),
                              function(x) ifelse(x < 0,
                                                 parse(text = paste0(-x, "^o", "*W")),
                                                 ifelse(x > 0,
                                                        parse(text = paste0(x, "^o", "*E")),
                                                        parse(text = paste0(x, "^o"))))
                              ))) +
  scale_y_continuous(breaks = seq(-90, 90, 45),
                     minor_breaks = seq(-90, 90, 15),
                     labels = unlist(
                       lapply(seq(-90, 90, 45),
                              function(x) ifelse(x < 0,
                                                 parse(text = paste0(-x, "^o", "*S")),
                                                 ifelse(x > 0,
                                                        parse(text = paste0(x, "^o", "*N")),
                                                        parse(text = paste0(x, "^o"))))
                              ))) +
  labs(title = "Title",
       subtitle = "Subtitle",
       caption = "Caption",
       tag = "Tag",
       x = "Longitude",
       y = "Latitude")
```

In the code below, a refactoring has been performed to extract the boilerplate
code for the country borders and axes scales. Notice that the code return a list
of items that include a `layer` object containing the borders and two `scale`
objects for the axes. The `+.gg` operator allows various types of plot
components to be added to a plot, including layers and scales. These can either
be added individually or a list of this can be added with a single `+` operator.

The function below extracts the common plot components into a single function
call that can be used as the basis for global maps. By doing so, this code can
then be used within a `ggplot()` call sequence to quickly add the borders and
scales that are common to all global maps. Effectively, creating a function like
this that returns a list of ggplot components acts like a macro. It is suggested
that any functions like this be prefixed with `mg_` by name to indicate they are
a macro ggplot operation.

The use and benefit of this is demonstrated in the following code. Notice the
structure of the function call that returns a `list`. Also note the conciseness
of the ggplot call sequence. As long as these macro calls return a list of
addable ggplot components, these calls can be placed anywhere inside the call
sequence.

```{r}
mg_global_map <- function() {
  return(list(
    borders(size = 0.1,
            colour = "black"),
    scale_x_continuous(breaks = seq(-180, 180, 90),
                       minor_breaks = seq(-180, 180, 30),
                       labels = unlist(
                         lapply(seq(-180, 180, 90),
                                function(x) ifelse(x < 0,
                                                   parse(text = paste0(-x, "^o", "*W")),
                                                   ifelse(x > 0,
                                                          parse(text = paste0(x, "^o", "*E")),
                                                          parse(text = paste0(x, "^o"))))
                                ))),
    scale_y_continuous(breaks = seq(-90, 90, 45),
                       minor_breaks = seq(-90, 90, 15),
                       labels = unlist(
                         lapply(seq(-90, 90, 45),
                                function(x) ifelse(x < 0,
                                                   parse(text = paste0(-x, "^o", "*S")),
                                                   ifelse(x > 0,
                                                          parse(text = paste0(x, "^o", "*N")),
                                                          parse(text = paste0(x, "^o"))))
                                )))
    ))
}

ggplot() +
  mg_global_map() +
  labs(title = "Title",
       subtitle = "Subtitle",
       caption = "Caption",
       tag = "Tag",
       x = "Longitude",
       y = "Latitude")  
```

### World Maps with `ggmap` Package

The code in this section demonstrates how global maps can be rendered using the
`ggmap` package. Not a lot of discussion is provided around the code examples
because of clear disadvantages in using this approach for current projects.

#### `ggmap` Risks

1. The package relies on Google and other third party mapping services.
2. The Google mapping services require an API key to be administered within a
   registered Google account. This key needs to be configured in the application
   and managed. Changing of the Google account or API key enabled within Google
   could render the application unusable requiring reconfiguring of the
   credentials.
3. Requiring a third party key for an API exposes the key owner to risk if the
   provider's terms-of-service change in any way.
4. The styling of political boundaries (i.e. countries, states, etc.) is not
   a trivial process when using third party mapping services. The maps lend
   themselves to data overlays but the political boundaries being displayed in
   the maps may not be available with certain providers.

#### `ggmap` Rewards

1. Map backgrounds such as terrains, satellite views, or roads are easily
   obtainable.
2. API provides access to advanced features such as reverse geocoding, travel
   routes, etc.
3. Automated labelling of geographical boundaries at various levels such as
   oceans, countries, etc.

```{r}
library(ggmap)
```

```{r}
world_map <- get_stamenmap(bbox = c(left = -179.99,
                                    bottom = -60.0,
                                    right = 180,
                                    top = 83),
                           maptype = "toner-lite",
                           zoom = 3,
                           force = TRUE,
                           crop = TRUE,
                           messaging = FALSE)
```

```{r}
ggmap(world_map)
```

## US State Maps

Maps of the United States of America can be derived from both the `maps::world`
and `maps::state` datasets. The data for the United States that exists in the
world dataset is indicated by the `region` variable being set to `USA`.
Collectively, these observations contain the political boundaries or borders of
the United States. It is important to remember the boundaries of the Unites
States not only include the 48 contiguous states on the North American
continent but also Alaska, Hawaii, and a number of other territories that are
not states.

The `maps` datasets contain a `subregion` variable that is used to create
subgroups within regions. For the US, it would seem that these groups would
translate to individual states and territories but not quite. There are a
number of observation in the US subset where there subregion is either numeric
(e.g. 73) or not indicated (i.e. `NA`). The code below shows the geographic
extents of each subregion indicating the minimum and maximim latitudes and
longitudes within each.

```{r}
map_data("world") %>%
  filter(region == "USA") %>%
  group_by(subregion) %>%
  summarise(count = n(),
            min_lat = min(lat),
            max_lat = max(lat),
            min_long = min(long),
            max_long = max(long)) %>%
  arrange(min_long)
```

Inspection of the subregions indicates that most pf the longitudes are negative
meaning they exist in the western hemisphere where the United States mainland
also exists. There are some parts of Alaska that extend past the $180^{\circ}$
line of longitide. Most longitude in Alaska are very negative but a few are
very positive ($179.78^{\circ} \text{E}$). For most US visualizations, grouping
will be done at the state level including all 50 states and possible Washington
D.C. Boundaries for the country can be collected by getting the US boundaries
that have negative longitudes from the world dataset. Boundaries for each of the
48 states and Washington D.C. can be obtained from the states dataset.

```{r}
map_us <- map_data("world") %>%
  filter(region == "USA" & long < 0)
map_us_48 <- map_data("state")
```

The national and state boundaries can then be plotted as a United States map
as follows.

```{r}
ggplot(mapping = aes(x = long, y = lat,
                     group = group)) +
  geom_polygon(data = map_us,
               color = "black",
               size = 0.1,
               fill = NA) +
  geom_polygon(data = map_us_48,
               color = "black",
               size = 0.1,
               fill = NA)
```

A macro function and example usage is presented below.

```{r}
mg_us_map <- function() {
  return(list(
  geom_polygon(data = map_us,
               mapping = aes(x = long, y = lat,
                     group = group),
               color = "black",
               size = 0.1,
               fill = NA),
  geom_polygon(data = map_us_48,
               mapping = aes(x = long, y = lat,
                     group = group),
               color = "black",
               size = 0.1,
               fill = NA),
  scale_x_continuous(breaks = seq(-180, -60, 30),
                     minor_breaks = seq(-180, -60, 15),
                     labels = unlist(
                       lapply(seq(-180, -60, 30),
                              function(x) ifelse(x < 0,
                                                 parse(text = paste0(-x, "^o", "*W")),
                                                 ifelse(x > 0,
                                                        parse(text = paste0(x, "^o", "*E")),
                                                        parse(text = paste0(x, "^o"))))
                              ))),
  scale_y_continuous(breaks = seq(10, 80, 10),
                     minor_breaks = seq(10, 80, 5),
                     labels = unlist(
                       lapply(seq(10, 80, 10),
                              function(x) ifelse(x < 0,
                                                 parse(text = paste0(-x, "^o", "*S")),
                                                 ifelse(x > 0,
                                                        parse(text = paste0(x, "^o", "*N")),
                                                        parse(text = paste0(x, "^o"))))
                              )))
  ))
}

ggplot() +
  mg_us_map() +
  labs(title = "Title",
       subtitle = "Subtitle",
       caption = "Caption",
       tag = "Tag",
       x = "Longitude",
       y = "Latitude")  
```
