---
title: "Geographic Mapping Techniques"
author:
- Jedidiah Harwood^[California State University - East Bay]
- Kurt Wydrinski^[California State University - East Bay]
date: "11/22/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(maps)
```

# Introduction

There are a number of techniques available for creating geographic maps within
R. Techniques using both the `maps`, `ggplot` and other `tidyverse` packages are
demonstrated in the following sections.

# Basic Techniques

## World Maps

The code below demonstrates the simplest world map. By default, borders are
drawn 1 unit thick in a medium grey color. Also notice that the map is centered
on the Prime Meridian at $0^{\circ}$ longitude.

```{r}
ggplot() +
  borders()
```

The simple map is just that, simple but it's not very appealing. Adding the
`size` graphic aesthetic parameter to the borders call, thinner border line can
be used in which improves the appeal of the map.

```{r}
ggplot() +
  borders(size = 0.1)    # thin borders
```

While the thin borders are an improvement, they become harder to see due to the
default medium-grey color. Using the `colour` parameter of `borders()` allows
the borders to be rendered in different colors. The code below demonstrates
using a thin, black border. This fine-line border looks cleaner and more refined
than the default border rendering.

```{r}
ggplot() +
  borders(size = 0.1,
          colour = "black")    # thin, black borders
```

The axes, titles, captions, etc. are either missing or not labelled very clearly
in the above plot. The following actions will address this:

1. Re-scale axes and label with proper units-of-measure.
2. Clearly label variables represented on the axes.
3. Add placeholder titling elements for further clarity. However, these are
   placeholders because it is expected that each world map data visualization
   will be appropriately titled for the information it is conveying. These
   guidelines should be followed when choosing appropriate titling elements:
   
   1. `title` - Indicate the main subject/point of data visualization (e.g.
      *Country Population*)
   2. `subtitle` - Indicate constraints, filters or transformations applied to
      the underlying data (e.g. *(by most recent reported census)*, *(in thousands of people)*,
      etc.)
   3. `caption` - **TODO: define standard**
   4. `tag` - **TODO: define standard**

The code below applies to actions to improve the ease in understanding the map.

```{r}
ggplot() +
  borders(size = 0.1,
          colour = "black") +    # thin, black borders
  scale_x_continuous(breaks = seq(-180, 180, 90),
                     minor_breaks = seq(-180, 180, 30),
                     labels = unlist(
                       lapply(seq(-180, 180, 90),
                              function(x) ifelse(x < 0,
                                                 parse(text = paste0(x, "^o", "*W")),
                                                 ifelse(x > 0,
                                                        parse(text = paste0(x, "^o", "*E")),
                                                        parse(text = paste0(x, "^o"))))
                              ))) +
  scale_y_continuous(breaks = seq(-90, 90, 45),
                     minor_breaks = seq(-90, 90, 15),
                     labels = unlist(
                       lapply(seq(-90, 90, 45),
                              function(x) ifelse(x < 0,
                                                 parse(text = paste0(x, "^o", "*S")),
                                                 ifelse(x > 0,
                                                        parse(text = paste0(x, "^o", "*N")),
                                                        parse(text = paste0(x, "^o"))))
                              ))) +
  labs(title = "Title",
       subtitle = "Subtitle",
       caption = "Caption",
       tag = "Tag",
       x = "Longitude",
       y = "Latitude")
```

In the code below, a refactoring has been performed to extract the boilerplate
code for the country borders and axes scales. Notice that the code return a list
of items that include a `layer` object containing the borders and two `scale`
objects for the axes. The `+.gg` operator allows various types of plot
components to be added to a plot, including layers and scales. These can either
be added individually or a list of this can be added with a single `+` operator.

The function below extracts the common plot components into a single function
call that can be used as the basis for global maps. By doing so, this code can
then be used within a `ggplot()` call sequence to quickly add the borders and
scales that are common to all global maps. Effectively, creating a function like
this that returns a list of ggplot components acts like a macro. It is suggested
that any functions like this be prefixed with `mg_` by name to indicate they are
a macro ggplot operation.

The use and benefit of this is demonstrated in the following code. Notice the
structure of the function call that returns a `list`. Also note the conciseness
of the ggplot call sequence. As long as these macro calls return a list of
addable ggplot components, these calls can be placed anywhere inside the call
sequence.

```{r}
mg_global_map <- function() {
  return(list(
    borders(size = 0.1,
            colour = "black"),
    scale_x_continuous(breaks = seq(-180, 180, 90),
                       minor_breaks = seq(-180, 180, 30),
                       labels = unlist(
                         lapply(seq(-180, 180, 90),
                                function(x) ifelse(x < 0,
                                                   parse(text = paste0(x, "^o", "*W")),
                                                   ifelse(x > 0,
                                                          parse(text = paste0(x, "^o", "*E")),
                                                          parse(text = paste0(x, "^o"))))
                                ))),
    scale_y_continuous(breaks = seq(-90, 90, 45),
                       minor_breaks = seq(-90, 90, 15),
                       labels = unlist(
                         lapply(seq(-90, 90, 45),
                                function(x) ifelse(x < 0,
                                                   parse(text = paste0(x, "^o", "*S")),
                                                   ifelse(x > 0,
                                                          parse(text = paste0(x, "^o", "*N")),
                                                          parse(text = paste0(x, "^o"))))
                                )))
    ))
}

ggplot() +
  mg_global_map() +
  labs(title = "Title",
       subtitle = "Subtitle",
       caption = "Caption",
       tag = "Tag",
       x = "Longitude",
       y = "Latitude")  
```
